Assessing the impacts of time to detection distribution assumptions on detection probability estimation

### 2017-07-13 Draft.tex
Remove comments

\documentclass[12pt]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{parskip}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage[margin=1cm]{caption}
%\usepackage{tabularx, booktabs}
%\usepackage{float}
\usepackage{adjustbox}
\usepackage[toc,page]{appendix}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{url}
\usepackage{xr}
\usepackage{lineno}

@@

\newif\ifdic   % Type \ifdic at the beginning of any optional section... use \fi to denote the end of the optional section
  % See: http://tex.stackexchange.com/questions/33576/conditional-typesetting-build
%\dictrue         % comment out if I want to hide DIC and model fit paragraphs

@@

\newcommand{\adam}[1]{{\color{blue} ADAM: #1}}
\newcommand{\jarad}[1]{{\color{Orange} JARAD: #1}}

@@

%\tableofcontents

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
The standard model for estimating detection from removal-sampled point-count surveys assumes that organisms at a survey site are detected at a constant rate; however, this assumption can often lead to biased estimates.  
%Detection rates can be influenced by organismal behaviors, survey methodology, and observer effort.  
%Failure to account for non-constant detection leads to biased estimates of detection and therefore of abundance.  
We consider a class of N-mixture models that allows for detection heterogeneity over time through a flexibly defined time-to-detection distribution (TTDD) and allows for fixed and random effects for both abundance and detection.
%We consider a class of N-mixture models that allow for detection heterogeniety through a mixture component, flexible two-parameter time-to-detection distributions (TTDDs), and allows for fixed and random effects for both abundance and detection.
%Whereas, under the standard approach, detection probability is modeled by dividing the observation period into equal-duration intervals, we instead model the detection {\it rate} in continuous time via a time-to-detection distribution (TTDD) embedded within a hierarchical N-mixture framework.  
Our model is thus a combination of survival time-to-event analysis with unknown-N, unknown-p abundance estimation.  
We specifically explore two-parameter families of TTDDs, e.g. gamma, that can additionally include a mixture component to model increased probability of detection in the initial observation period.
Based on simulation analyses, we find that modeling a TTDD by using a two-parameter family is necessary when data have a chance of arising from a distribution of this nature.
In addition, models with a mixture component can outperform non-mixture models even when the truth is non-mixture.  
Finally, we analyze an Overbird data set from the Chippewa National Forest using mixed effect models for both abundance and detection.
We demonstrate that the effects of explanatory variables on abundance and detection are consistent across mixture TTDDs but that flexible TTDDs result in lower estimated probabilities of detection and therefore higher estimates of abundance. 
%We find that modeling a TTDD by using a mixture component for increased probability of detection in the initial observation period is necessary when data have a chance of arising from distributions of this nature.
%In addition, the flexibility offered by two-parameter distributions, e.g. gamma, can outperform exponential based models when the truth is exponential or non-exponential.
%Finally, we analyze an Overbird data set from the Chippewa National Forest using mixed effect models for both abundance and detection and demonstrate that the effects of explanatory variables on abundance and detection are consistent across mixture TTDDs and that flexible TTDDs result in lower estimated probabilities of detection and therefore higher estimates of abundance. 
% We apply this model to Ovenbird counts from the Chippewa National Forest and to datasets simulated under different time-to-detection patterns.  
% Models assuming constant detection rates produce biased estimates of detection when true detection rates vary with time, whereas models allowing for variable detection (assuming gamma, Weibull, or lognormal distributed times to detection) produce less biased estimates of the detection probability and nominal credible interval coverage.  
% Models ignoring detection heterogeneity across subgroups yield biased estimates of detection when such heterogeneity exists, whereas models accounting for detection heterogeneity (modeled as a mixture) return reasonable coverage rates and can outperform heterogeneity-ignorant models even when there is no heterogeneity.

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
Removal sampling, where individuals are solely counted on their first capture, provides one established methodology for estimating detection probabilities \citep{Farnsworth2002}.
% cost-effective in that it requires neither multiple observers nor multiple site visits.
%Removal sampling is a species abundance surveying methodology in which observers capture and physically remove animals at a survey site over a series of trapping sessions.
%Assuming that individual capture rates remain constant over all trapping sessions, it is possible to estimate the proportion of individuals still not captured from the pattern of captures over time .  
%\citet{Farnsworth2002} adapted removal sampling to point-count surveys, suggesting that if: (i) observers record only the first time-to-detection for each animal, and (ii) times to first detection are interval-censored, then data will be similar in for to traditional removal sampled data.
%In the analysis of such data, the observation period is subdivided into unit intervals of equal duration, each of which is equivalent to a trapping session.
%Farnsworth's removal assumptions:\\
%(i) Closed population during survey\\
%(ii) no double-counting\\
%(iii) easy-to-detect group is entirely counted during first interval\\
%(iv) TTDD for hard-to-detect group is correct after the end of the observation period\\
%(v) if using limited radius counts, they are accurate
% Johnson2008: all birds present are available for detection
A standard assumption in removal sampling is a constant detection rate throughout the observation period, but this assumption is often unjustified \citep{Alldredge2007, LeeMarsden2008}.  
In particular, animal behaviors such as intermittent singing in birds and frogs or diving in whales \citep{Scott2005, Diefenbach2007, Reidy2011}, differences in behavior across subgroups of animals \citep{Otis1978, Farnsworth2002}, observer impacts on animal behaviors \citep{McSheaRappole1997, Rosenstock2002, Alldredge2007}, and variations in observer effort, e.g. saturation or lack of settling down period \citep{Petit1995, LeeMarsden2008, Johnson2008}, can all lead to time-varying rates of detection.

@@

%Application of this methodology generally assumes that organisms at a survey site are detected at a constant rate \citep{Farnsworth2005, Etterson2009, Solymos2013, Amundson2014, Reidy2016}; however, this assumption is often not justified in the field.  
%Failure to account for non-constant detection leads to biased estimates of detection and therefore of abundance.

@@

% \adam{Based on further lit review, I think I need to insert a paragraph here about what time-varying modeling has been done.  
% In the fisheries world, models of time-varying catchability have been implemented -- see Schnute 1983, Scruton and Gibson (ref: Mantyniemi), and Mantyniemi2015 for examples.  
% Also, Alldredge 2007 implements a complete-history of detection model (technically mark-recapture, not removal) that estimates interval-specific detection probabilities (complete with covariates and detection heterogeneity).  
% However, it cannot be applied in a removal context, because removal data contains only first times to detection, and the Alldredge model requires subsequent observations -- removal models compensate by specifying a TTDD, thus enabling us to extrapolate beyond the observation period.}

Section \ref{sec:data} provides a description of the interval-censored time-to-detection avian point count data under consideration.
Section \ref{sec:model} introduces an N-mixture model with a generically defined TTDD for estimating abundance from removal-sampled point-count surveys.
%In addition, we discuss reasonable default priors and MCMC analysis used to estimate parameters in these models.
Section \ref{sec:sim} provides three simulation studies to assess the impact of TTDD choice on estimated detection probability. 
Section \ref{sec:ovenbirds} analyzes an Ovenbird data set under different TTDDs to determine the impact of this choice on estimated detection probability and therefore estimated abundance.

@@

Our analysis is motivated by avian point-count surveys in Chippewa National Forest from 2008-2013 as part of the Minnesota Forest Breeding Bird Project (MNFB) \citep{Hanowski1995}.  
%However, the model we describe is appropriate for point-count surveys of any species, so long as time-to-first-detection data are available.  
For our analysis, we focus on Ovenbird counts selected from one habitat type: sawtimber red pine stands with no recent logging activity.  
Each stand had up to four sites with sufficient geographical distance between sites to reduce or eliminate overlapping territories.
This dataset includes 947 Ovenbirds counted in 381 surveys at 65 sites with site-specific variables including site age,  stock density, and an indicator of select-/partial-cut logging during the 1990s.
% The counts of Ovenbirds by interval, across all sites, were 596, 69, 62, 68, 43, 33, 35, 27, and 14.

@@

Suppose that, for each survey $s$ ($s=1,\dots,S$), $N_s$ individuals are present.  
Imagine an observer could remain at the survey location until every individual is detected, recording the time to detection $t_{sb}$ (for bird, $b=1,\dots,N_s$) for each.
Assuming detection times for all individuals at a survey are independent, identically distributed according to a common time-to-detection distribution (TTDD), we define $T_{sb}$ as a random variable with cumulative distribution function (cdf) $F_T(t)$ and probability density function (pdf) $f_T(t)$.
In practice, times to first detection are truncated due to a finite survey length of $C$, meaning each individual has detection probability $\pdet=F_T(C)$.  
The conditional distribution of observed detection times consequently has pdf $\ftdt= f_T(t)/F_T(C)$ for $0<t<C$, cdf $F_{T|\dt}(t|\dt) = \int_0^t f_{T|\dt}(x|\dt) dx$, and instantaneous detection rate, or hazard function, $h(t) = f_T(t) / [1-F_T(t)]$. 
%$T_{sb|\dt} \ind f_{T|\dt}(t_{sb}|\dt)$ where $T_{sb|\dt}$ is observed only for those $\nobs_s$ individuals whose observed time to first detection is less than $C$, i.e. $t_{sb}<C$.

@@

%We model the number of individuals at survey $s$ for which $t_{sb}$ is actually observed as $\nobs_s \ind \mbox{Binomial}\left(N_{s}, \pdet\right)$. 
%Using maximum likelihood, we can estimate the parameters of a TTDD from the observed times to first detection and thereby estimate $\pdet$ and its uncertainty.
%With an estimate and uncertainty for $\pdet$, we are in a scenario of a binomial model with unknown $N_s$ and ``known'' $p$ and thus can estimate $N_s$.
%In these models, point estimates of $N_s$ are unstable unless $\pdet>0.41$ \citep{Olkin1981}.
%One approach to regularizing these estimates is to construct a hierarchical model for site-specific abundance, e.g. $N_s\ind \Po(\lambda)$ \citep{Raftery1988, Royle2004NMixture}. 
%Assuming Poisson abundance, we can decompose $N_{s}$ into observed and unobserved portions: $\nobs \sim Po(\lambda \pdet)$ and, independently, $n_{s}^{(unobs)} \sim \Po\left(\lambda[1-\pdet]\right)$.
%Although alternative distributions could be considered, e.g. negative binomial, our experience with Ovenbird point counts suggests that, after accounting for appropriate explanatory variables, the resulting abundances are likely underdispersed rather than overdispersed, and thus we will use the Poisson assumption here.

@@

We fit the models by MCMC sampling using the Bayesian statistical software Stan, implemented via the R package \texttt{rstan} version 2.8.0 \citep{Rstan2016}.  
%Stan model code and \texttt{rstan} code to run the models can be found in the Supplementary Materials.  
We discarded half of the iterations as warmup and thinned by 10.  
We monitored convergence of the MCMC chains using Geweke z-score diagnostics \citep{Geweke1991} and reran models if lack of convergence was indicated by a non-normal distribution of the z-scores or if the effective sample size for any parameter was below 1000.  
The number of iterations used depended on the model and is detailed later. 
For most models, we accepted Stan defaults for initial values; however, gamma and Weibull models sometimes failed to run unless care was taken in the specification of initial values.

%Thus, the link functions in our detection model are:
%\begin{alignat}{3}
%&\text{Exponential:\;} &&\log(\varphi_{s}) &&= \text{Intercept}^D + \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D = -\log(E[T_{sq}])\\
%&\text{Gamma:} &&\log(\varphi_{s}) &&= \text{Intercept}^D + \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D = -\log(E[T_{sq}]) + \log(\alpha)\\
%&\text{Weibull:}  &&\log(\varphi_{s}) &&= \text{Intercept}^D + \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D = -\log(E[T_{sq}]) + \log\left(\Gamma\left(1 + 1/k\right)\right)\\
%&\text{Lognormal:} &&-\mu_{s} &&= \text{Intercept}^D + \vX_{s}^D\vbeta^D + \vZ_{s}^D\vxi^D = -\log(E[T_{sq}]) + \sigma_{det}^2/2
%\end{alignat}

%\subsection{Abundance Model}

% We distinguish between non-constant detection and detection heterogeneity, which can be modeled in similar ways but refer to different mechanisms.  
% We reserve `detection heterogeneity' for differences across population subgroups.  
% It is often modeled using a mixture of detection distributions.  
% \citet{Farnsworth2002} introduce a finite-mixture for detection probabilities distinguishing between easy-to-detect individuals, which are always detected in first observation interval, and hard-to-detect individuals, which are detected at a constant rate.  
%  Heterogeneity at the individual level can be modeled as random effects in the calculation of detection probabilities \citep{DorazioRoyle2003, Mantyniemi2005}.  
% We use `non-constant detection' when detection rates for an organism change with time.  
% Models constructed to address detection heterogeneity may provide reasonable estimates for non-constant detection scenarios and vice versa \citep{Mantyniemi2005}.

%\adam{Individual effects on detection might suffer from identifiability issues. This was the subject of a dust-up between Pledger and Dorazio/Royle from 2003-2005.}

\ifdic\
\subsection{Model diagnostics}

To assess goodness of fit, we relied on posterior predictive checks \citep{Gelman1996}.  
%Posterior predictive checks assess whether a fitted model can effectively replicate datasets that look like the original data in key aspects, which are measured by check statistics.  
In particular, we are concerned with the goodness of fit for the TTDD distribution, so we defined a check statistic for each observation interval:
\[D(n_i) = \sum\limits_{s} n_{si} \big/ \sum\limits_{si} n_{si}\]
which is the proportion of the total count that occurs during interval $i$ marginally across all surveys.  
The posterior predictive p-value associated with each $D(n_i)$ is the proportion of replicate check statistics that have a larger value than the check statistic for the actual data.  
Posterior predictive p-values near to 0 or 1 are potentially indicators of a misspecified model.
We simulated replicate datasets and calculated the check statistic at each iteration of our MCMC sampling.

We focus inference on the overall detection probability across all surveys: $\pdet = \sum\limits_{s}\nobs_s \big/$ $\sum\limits_{s}N_{s}$.  
Within simulation studies, we compared model estimates to true $\pdet$ values by percent coverage and by posterior p-values --- the proportion of MCMC samples having a smaller value than the true value.

We compared different models by means of their goodness of fit and deviance information criterion (DIC) \citep{Spiegelhalter2002}.  
For the full model with random effects (Ovenbirds and Sim3), which is a missing data model, we used formulation DIC$_7$ from \citet{Celeux2006} for its ease of computation, but we acknowledge that it suffers some difficulties \citep{Celeux2006, Li2014}.
\fi

@@

% FIRST SUBMISSION TABLE
%\ifdic
%\begin{table}[ht]\centering\small
%\begin{tabular}{l|l|l|l|cccc|cccc||r}
% \multicolumn{4}{c|}{ } & \multicolumn{4}{c|}{\underline{Non-mixture model}} & \multicolumn{4}{c||}{\underline{Mixture model}} & \\
% \multicolumn{4}{c|}{ } & Med. $p$ & Q($p$) & 50\% & 90\% & Med. $p$ & Q($p$) & 50\% & 90\% & $\Delta$ DIC \\ 
%  \hline
%  \hline
% \parbox[t]{2mm}{\multirow{14}{*}{\rotatebox[origin=c]{90}{TTDD used to simulate data}}} & \parbox[t]{2mm}{\multirow{7}{*}{\rotatebox[origin=c]{90}{Non-mixture}}} & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Nonpk.}}} & Gamma & 0.76 & 0.41 & 0.75 & 0.88 & 0.84 & 0.66 & 0.56 & 0.94 & 0.24 \\ 
% & & &   Lognormal & 0.66 & 0.17 & 0.31 & 0.62 & 0.78 & 0.48 & 0.62 & 1.00 & -0.67 \\ 
% & & &   Weibull & 0.69 & 0.25 & 0.38 & 0.75 & 0.79 & 0.51 & 0.75 & 1.00 & 0.37 \\ 
%  \cline{3-13}
% & & &   Exponential & 0.80 & 0.54 & 0.44 & 0.94 & 0.79 & 0.38 & 0.31 & 0.88 & 1.48 \\ 
%  \cline{3-13}
% & & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Peaked}}} &  Gamma & 0.80 & 0.55 & 0.50 & 1.00 & 0.82 & 0.70 & 0.38 & 0.88 & 0.96 \\ 
% & & &   Lognormal & 0.78 & 0.31 & 0.50 & 0.94 & 0.80 & 0.48 & 0.62 & 1.00 & 0.97 \\ 
% & & &   Weibull & 0.79 & 0.49 & 0.56 & 0.94 & 0.82 & 0.66 & 0.44 & 0.88 & 1.28 \\ 
%  \cline{2-13}
%& \parbox[t]{2mm}{\multirow{7}{*}{\rotatebox[origin=c]{90}{Mixture}}} & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Nonpk.}}} & Gamma & 0.67 & 0.17 & 0.12 & 0.81 & 0.76 & 0.41 & 0.69 & 1.00 & 0.51 \\ 
% & & &   Lognormal & 0.56 & 0.04 & 0.00 & 0.31 & 0.72 & 0.30 & 0.44 & 0.88 & 0.33 \\ 
% & & &   Weibull & 0.51 & 0.02 & 0.00 & 0.06 & 0.71 & 0.32 & 0.44 & 1.00 & 2.40 \\ 
%  \cline{3-13}
% & & &   Exponential & 0.96 & 1.00 & 0.00 & 0.00 & 0.77 & 0.37 & 0.38 & 0.94 & 122.81 \\ 
%  \cline{3-13}
% & & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Peaked}}} & Gamma & 0.28 & 0.00 & 0.00 & 0.00 & 0.74 & 0.33 & 0.31 & 0.88 & 31.84 \\ 
% & & &   Lognormal & 0.22 & 0.00 & 0.00 & 0.00 & 0.76 & 0.36 & 0.38 & 0.94 & 63.65 \\ 
% & & &   Weibull & 0.22 & 0.00 & 0.00 & 0.00 & 0.70 & 0.29 & 0.56 & 0.94 & 28.30 \\ 
%   \hline
%\end{tabular}
%\caption{Summary of mixture vs. non-mixture model fits.  
%In all cases, the inference model family matches the dataset family.  
%Med $p$: average across simulations of the posterior median of $\pdet$ (true value = 0.80).  
%Q($p$): average proportion of the posterior distribution of $\pdet$ that is larger than the true value.  
%50\% and 90\% coverage is expressed as the proportion of 16 simulations for which the true value of $\pdet$ lies within the appropriate credible interval.  
%$\Delta$ DIC: mean difference in DIC between the incorrect-mixture model minus the true model.
%}
%\label{tbl:sim1}
%\end{table}
%
%\else
%
%\begin{table}[ht]\centering\small
%\begin{tabular}{l|l|l|l|cccc|cccc}
% \multicolumn{4}{c|}{ } & \multicolumn{4}{c|}{\underline{Non-mixture model}} & \multicolumn{4}{c}{\underline{Mixture model}} \\
% \multicolumn{4}{c|}{ } & Med. $p$ & Q($p$) & 50\% & 90\% & Med. $p$ & Q($p$) & 50\% & 90\% \\ 
%  \hline
%  \hline
% \parbox[t]{2mm}{\multirow{14}{*}{\rotatebox[origin=c]{90}{TTDD used to simulate data}}} & \parbox[t]{2mm}{\multirow{7}{*}{\rotatebox[origin=c]{90}{Non-mixture}}} & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Nonpk.}}} & Gamma & 0.76 & 0.41 & 0.75 & 0.88 & 0.84 & 0.66 & 0.56 & 0.94 \\ 
% & & &   Lognormal & 0.66 & 0.17 & 0.31 & 0.62 & 0.78 & 0.48 & 0.62 & 1.00  \\ 
% & & &   Weibull & 0.69 & 0.25 & 0.38 & 0.75 & 0.79 & 0.51 & 0.75 & 1.00 \\ 
%  \cline{3-12}
% & & &   Exponential & 0.80 & 0.54 & 0.44 & 0.94 & 0.79 & 0.38 & 0.31 & 0.88  \\ 
%  \cline{3-12}
% & & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Peaked}}} &  Gamma & 0.80 & 0.55 & 0.50 & 1.00 & 0.82 & 0.70 & 0.38 & 0.88 \\ 
% & & &   Lognormal & 0.78 & 0.31 & 0.50 & 0.94 & 0.80 & 0.48 & 0.62 & 1.00 \\ 
% & & &   Weibull & 0.79 & 0.49 & 0.56 & 0.94 & 0.82 & 0.66 & 0.44 & 0.88 \\ 
%  \cline{2-12}
%& \parbox[t]{2mm}{\multirow{7}{*}{\rotatebox[origin=c]{90}{Mixture}}} & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Nonpk.}}} & Gamma & 0.67 & 0.17 & 0.12 & 0.81 & 0.76 & 0.41 & 0.69 & 1.00 \\ 
% & & &   Lognormal & 0.56 & 0.04 & 0.00 & 0.31 & 0.72 & 0.30 & 0.44 & 0.88 \\ 
% & & &   Weibull & 0.51 & 0.02 & 0.00 & 0.06 & 0.71 & 0.32 & 0.44 & 1.00 \\ 
%  \cline{3-12}
% & & &   Exponential & 0.96 & 1.00 & 0.00 & 0.00 & 0.77 & 0.37 & 0.38 & 0.94 \\ 
%  \cline{3-12}
% & & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Peaked}}} & Gamma & 0.28 & 0.00 & 0.00 & 0.00 & 0.74 & 0.33 & 0.31 & 0.88 \\ 
% & & &   Lognormal & 0.22 & 0.00 & 0.00 & 0.00 & 0.76 & 0.36 & 0.38 & 0.94 \\ 
% & & &   Weibull & 0.22 & 0.00 & 0.00 & 0.00 & 0.70 & 0.29 & 0.56 & 0.94 \\ 
%   \hline
%\end{tabular}
%\caption{Summary of mixture vs. non-mixture model fits.  
%In all cases, the inference model family matches the dataset family.  
%Med $p$: average across simulations of the posterior median of $\pdet$ (true value = 0.80).  
%Q($p$): average proportion of the posterior distribution of $\pdet$ that is larger than the true value.  
%50\% and 90\% coverage is expressed as the proportion of 16 simulations for which the true value of $\pdet$ lies within the appropriate credible interval.  
%}
%\label{tbl:sim1}
%\end{table}
%\fi

% THIS IS SIMILAR TO THE ABOVE TABLE FOR ALL FOUR PDET SIMULATIONS.
% WE DEEMED IT TO OPPRESSIVE AND INSTEAD GENERATED THE FIGURES CURRENTLY IN THE PLOT
%\begin{table}[ht]\centering\footnotesize
%\setlength{\tabcolsep}{4pt} %default is 6
%\begin{tabular}{l|l|l|l|cccc|cccc||cccc|cccc}
% \multicolumn{4}{c|}{ } & \multicolumn{8}{c}{\underline{Non-mixture model}} & \multicolumn{8}{c}{\underline{Mixture model}} \\
% \multicolumn{4}{c|}{ } & \multicolumn{4}{c}{\underline{Posterior median}} & \multicolumn{4}{c}{\underline{Coverage}} & \multicolumn{4}{c}{\underline{Posterior median}} & \multicolumn{4}{c}{\underline{Coverage}}\\
% \multicolumn{4}{c|}{ } & p50 & p65 & p80 & p95 & p50 & p65 & p80 & p95 & p50 & p65 & p80 & p95 & p50 & p65 & p80 & p95 \\
%  \hline
%  \hline
%\parbox[t]{2mm}{\multirow{14}{*}{\rotatebox[origin=c]{90}{TTDD used to simulate data}}} & \parbox[t]{2mm}{\multirow{7}{*}{\rotatebox[origin=c]{90}{Non-mixture}}} & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Nonpk.}}} & Gamma & 0.67 & 0.69 & 0.75 & 0.95 & 0.11 & 0.78 & 0.45 & 0.59 & 0.76 & 0.78 & 0.83 & 0.96 & 0.00 & 0.35 & 0.55 & 0.61 \\ 
%   &  &  & Lognormal & 0.54 & 0.59 & 0.71 & 0.94 & 0.81 & 0.45 & 0.29 & 0.47 & 0.68 & 0.73 & 0.82 & 0.96 & 0.05 & 0.63 & 0.63 & 0.44 \\ 
%   &  &  & Weibull & 0.55 & 0.60 & 0.72 & 0.94 & 0.79 & 0.59 & 0.41 & 0.43 & 0.69 & 0.73 & 0.82 & 0.96 & 0.16 & 0.59 & 0.59 & 0.52 \\ 
%\cline{3-20}
%   &  &  & Exponential & 0.49 & 0.64 & 0.80 & 0.95 & 0.51 & 0.50 & 0.54 & 0.48 & 0.45 & 0.61 & 0.78 & 0.94 & 0.47 & 0.33 & 0.48 & 0.42 \\ 
%\cline{3-20}
%   & & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Peaked}}} & Gamma & 0.48 & 0.64 & 0.79 & 0.95 & 0.57 & 0.46 & 0.45 & 0.44 & 0.57 & 0.68 & 0.81 & 0.95 & 0.42 & 0.51 & 0.45 & 0.40 \\ 
%   &  &  & Lognormal & 0.49 & 0.64 & 0.80 & 0.95 & 0.53 & 0.36 & 0.52 & 0.44 & 0.54 & 0.67 & 0.81 & 0.95 & 0.37 & 0.44 & 0.53 & 0.44 \\ 
%   &  &  & Weibull & 0.48 & 0.61 & 0.78 & 0.95 & 0.40 & 0.46 & 0.38 & 0.49 & 0.59 & 0.69 & 0.81 & 0.96 & 0.37 & 0.46 & 0.49 & 0.42 \\ 
%\cline{2-20}
%\cline{2-20}
%& \parbox[t]{2mm}{\multirow{7}{*}{\rotatebox[origin=c]{90}{Mixture}}} & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Nonpk.}}} & Gamma & 0.62 & 0.60 & 0.64 & 0.85 & 0.42 & 0.78 & 0.11 & 0.03 & 0.71 & 0.71 & 0.75 & 0.91 & 0.10 & 0.81 & 0.64 & 0.31 \\ 
%   &  &  & Lognormal & 0.50 & 0.52 & 0.58 & 0.85 & 0.95 & 0.12 & 0.04 & 0.03 & 0.64 & 0.67 & 0.74 & 0.92 & 0.09 & 0.93 & 0.51 & 0.38 \\ 
%   &  &  & Weibull & 0.50 & 0.51 & 0.54 & 0.71 & 0.92 & 0.23 & 0.01 & 0.01 & 0.64 & 0.66 & 0.70 & 0.87 & 0.39 & 0.92 & 0.46 & 0.28 \\ 
%\cline{3-20}
%   &  &  & Exponential & 0.88 & 0.93 & 0.96 & 0.99 & 0.00 & 0.00 & 0.00 & 0.00 & 0.50 & 0.63 & 0.78 & 0.94 & 0.54 & 0.45 & 0.42 & 0.51 \\ 
%\cline{3-20}
% & & \parbox[t]{2mm}{\multirow{3}{*}{\rotatebox[origin=c]{90}{Peaked}}} & Gamma & 0.24 & 0.26 & 0.28 & 0.37 & 0.00 & 0.00 & 0.00 & 0.00 & 0.52 & 0.58 & 0.72 & 0.94 & 0.76 & 0.51 & 0.40 & 0.41 \\ 
%   &  &  & Lognormal & 0.18 & 0.20 & 0.22 & 0.27 & 0.00 & 0.00 & 0.00 & 0.00 & 0.49 & 0.61 & 0.78 & 0.94 & 0.66 & 0.58 & 0.46 & 0.43 \\ 
%   &  &  & Weibull & 0.18 & 0.19 & 0.22 & 0.28 & 0.00 & 0.00 & 0.00 & 0.00 & 0.52 & 0.56 & 0.69 & 0.93 & 0.80 & 0.44 & 0.33 & 0.47 \\ 
%  \hline
%\end{tabular}
%\caption{Summary of mixture vs. non-mixture model fits over 100 replicates.
%In all cases, the inference model family matches the dataset family.  
%Column titles denote the true detection probability (e.g. `p80' means $\pdet = 0.80$). 
%Medians are averaged across all replicates.
%Coverage is for 50\% credible intervals.
%For nominal coverage, we expect values between (0.41, 0.59) 95\% of the time.}
%\label{tbl:sim1}
%\end{table}

% Median posterior abundance from mixture models was unbiased for non-mixture data and positively biased by only 4-15\% for mixture data.  
% Median posterior abundance from non-mixture models applied to non-mixture data were unbiased (exponential and peaked) or biased by 5-24\% (nonpeaked); however, for mixture data, the non-mixture models were biased by 20-60\% (nonpeaked), -17\% (exponential), and 183-265\% (peaked).  
% For the time-varying distributions, non-mixture models not only had greater bias, but 95\% credible intervals were generally 10-40\% wider (470\% wider for peaked mixture data).
% For exponential datasets, the non-mixture model credible intervals were 28\% narrower for non-mixture data and an unrealistic 95\% narrower for mixture data.

% Possible other comments
%(1) For nonpeaked data, the true inference model is biased low\\
%(2) Mixture inference models generate higher estimates than non-mixture models\\
%(3) True models are generally good\\
%(4) Estimates of non-mixture peaked data are much more precise\\
%(5) Note that plots of TTDD at the median are indistinguishable from the truth except in the blatant fail scenario

@@

% FIRST SUBMISSION TABLE
%\ifdic
%\begin{table}[ht]
%\footnotesize\centering
%\begin{tabular}{l|l|l|ccccc|ccccc}

>>>>

\documentclass[12pt]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{parskip}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames]{xcolor}
\usepackage[margin=1cm]{caption}
\usepackage{adjustbox}
\usepackage[toc,page]{appendix}
\usepackage{enumitem}
\usepackage{multirow}
\usepackage{setspace}
\usepackage{url}
\usepackage{xr}
\usepackage{lineno}

@@

@@

@@

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
The standard model for estimating detection from removal-sampled point-count surveys assumes that organisms at a survey site are detected at a constant rate; however, this assumption can often lead to biased estimates.  
We consider a class of N-mixture models that allows for detection heterogeneity over time through a flexibly defined time-to-detection distribution (TTDD) and allows for fixed and random effects for both abundance and detection.
Our model is thus a combination of survival time-to-event analysis with unknown-N, unknown-p abundance estimation.  
We specifically explore two-parameter families of TTDDs, e.g. gamma, that can additionally include a mixture component to model increased probability of detection in the initial observation period.
Based on simulation analyses, we find that modeling a TTDD by using a two-parameter family is necessary when data have a chance of arising from a distribution of this nature.
In addition, models with a mixture component can outperform non-mixture models even when the truth is non-mixture.  
Finally, we analyze an Overbird data set from the Chippewa National Forest using mixed effect models for both abundance and detection.
We demonstrate that the effects of explanatory variables on abundance and detection are consistent across mixture TTDDs but that flexible TTDDs result in lower estimated probabilities of detection and therefore higher estimates of abundance.

@@

Abundance estimates from animal point-count surveys require accurate estimates of detection probabilities.  
Removal sampling, where individuals are solely counted on their first capture, provides one established methodology for estimating detection probabilities \citep{Farnsworth2002}.
A standard assumption in removal sampling is a constant detection rate throughout the observation period, but this assumption is often unjustified \citep{Alldredge2007, LeeMarsden2008}.  
In particular, animal behaviors such as intermittent singing in birds and frogs or diving in whales \citep{Scott2005, Diefenbach2007, Reidy2011}, differences in behavior across subgroups of animals \citep{Otis1978, Farnsworth2002}, observer impacts on animal behaviors \citep{McSheaRappole1997, Rosenstock2002, Alldredge2007}, and variations in observer effort, e.g. saturation or lack of settling down period \citep{Petit1995, LeeMarsden2008, Johnson2008}, can all lead to time-varying rates of detection.

@@

@@

Section \ref{sec:data} provides a description of the interval-censored time-to-detection avian point count data under consideration.
Section \ref{sec:model} introduces an N-mixture model with a generically defined TTDD for estimating abundance from removal-sampled point-count surveys.
Section \ref{sec:sim} provides three simulation studies to assess the impact of TTDD choice on estimated detection probability. 
Section \ref{sec:ovenbirds} analyzes an Ovenbird data set under different TTDDs to determine the impact of this choice on estimated detection probability and therefore estimated abundance.

@@

Our analysis is motivated by avian point-count surveys in Chippewa National Forest from 2008-2013 as part of the Minnesota Forest Breeding Bird Project (MNFB) \citep{Hanowski1995}.  
For our analysis, we focus on Ovenbird counts selected from one habitat type: sawtimber red pine stands with no recent logging activity.  
Each stand had up to four sites with sufficient geographical distance between sites to reduce or eliminate overlapping territories.
This dataset includes 947 Ovenbirds counted in 381 surveys at 65 sites with site-specific variables including site age,  stock density, and an indicator of select-/partial-cut logging during the 1990s.

@@

Suppose that, for each survey $s$ ($s=1,\dots,S$), $N_s$ individuals are present.  
Imagine an observer could remain at the survey location until every individual is detected, recording the time to detection $t_{sb}$ (for bird, $b=1,\dots,N_s$) for each.
Assuming detection times for all individuals at a survey are independent, identically distributed according to a common time-to-detection distribution (TTDD), we define $T_{sb}$ as a random variable with cumulative distribution function (cdf) $F_T(t)$ and probability density function (pdf) $f_T(t)$.
In practice, times to first detection are truncated due to a finite survey length of $C$, meaning each individual has detection probability $\pdet=F_T(C)$.  
The conditional distribution of observed detection times consequently has pdf $\ftdt= f_T(t)/F_T(C)$ for $0<t<C$, cdf $F_{T|\dt}(t|\dt) = \int_0^t f_{T|\dt}(x|\dt) dx$, and instantaneous detection rate, or hazard function, $h(t) = f_T(t) / [1-F_T(t)]$.

@@

@@

We fit the models by MCMC sampling using the Bayesian statistical software Stan, implemented via the R package \texttt{rstan} version 2.8.0 \citep{Rstan2016}.  
We discarded half of the iterations as warmup and thinned by 10.  
We monitored convergence of the MCMC chains using Geweke z-score diagnostics \citep{Geweke1991} and reran models if lack of convergence was indicated by a non-normal distribution of the z-scores or if the effective sample size for any parameter was below 1000.  
The number of iterations used depended on the model and is detailed later. 
For most models, we accepted Stan defaults for initial values; however, gamma and Weibull models sometimes failed to run unless care was taken in the specification of initial values.

@@

@@
